---
title: "webnovel"
author: "author"
format: pdf
editor: visual
---

```{r}
combined_all_df <- read_csv("data/combined_qidian_webnovel_paragraphs_comments_sentiment.csv")
```

```{r}
count_comments_ratio <- combined_all_df |>
  mutate(has_CN_comment = ifelse(!is.na(avg_comm_sentiment_CN), 1, 0),
         has_EN_comment = ifelse(!is.na(avg_comm_sentiment_EN), 1, 0)) |>
  summarize(ratio_CN = round(mean(has_CN_comment), 2),
            ratio_EN = round(mean(has_EN_comment), 2))

count_comments_ratio
```

```{r message=FALSE}
long_df <- combined_all_df |>
  pivot_longer(c(par_sentiment_CN, par_sentiment_EN),
               names_to = "source",
               values_to = "par_sentiment") |>
  pivot_longer(c(avg_comm_sentiment_CN, avg_comm_sentiment_EN),
               names_to = "source_comm_sentiment",
               values_to = "avg_comm_sentiment") |>
  mutate(source = case_when(source == "par_sentiment_CN" ~ "Qidian",
                            source == "par_sentiment_EN" ~ "WebNovel"),
         source_comm_sentiment = case_when(source_comm_sentiment == "avg_comm_sentiment_CN" ~ "Qidian",
                                          source_comm_sentiment == "avg_comm_sentiment_EN" ~ "WebNovel")) |>
  group_by(bookId, chapter_index, paragraph_index) |>
  filter(source == source_comm_sentiment) |>
  ungroup() |>
  select(-source_comm_sentiment) |>
  relocate(bookId, new_index, chapter_index, paragraph_index, percentage_index,
           source, par_sentiment, avg_comm_sentiment)
```

Let's smooth the sentiment using a moving window of 10 or 50 paragraphs.

```{r message=FALSE}
# This is the original code used by Pianzola et al., 2021, but it doesn't handle NAs well enough for cases with sparse comments

# library(zoo) # for rolling mean

# long_df_roll_zoo <- long_df |>
#   group_by(bookId, source) |>
#   mutate(par_sentiment_roll_11 = rollmean(par_sentiment, 10, fill = NA),
#          par_sentiment_roll_51 = rollmean(par_sentiment, 50, fill = NA),
#          comm_sentiment_roll_11 = rollmean(avg_comm_sentiment, 10, fill = NA),
#          comm_sentiment_roll_51 = rollmean(avg_comm_sentiment, 50, fill = NA)) |>
#   ungroup()
```

```{r}
library(slider) # for rolling mean with NA handling

long_df_roll <- long_df |>
  group_by(bookId, source) |>
  mutate(
    par_sentiment_roll_11 = slider::slide_dbl(par_sentiment,
      ~ {v <- mean(.x, na.rm = TRUE); if (is.nan(v)) NA_real_ else v},
      .before = 5, .after = 5, .complete = TRUE),
    par_sentiment_roll_51 = slider::slide_dbl(par_sentiment,
      ~ {v <- mean(.x, na.rm = TRUE); if (is.nan(v)) NA_real_ else v},
      .before = 25, .after = 25, .complete = TRUE),
    comm_sentiment_roll_11 = slider::slide_dbl(avg_comm_sentiment,
      ~ {v <- mean(.x, na.rm = TRUE); if (is.nan(v)) NA_real_ else v},
      .before = 5, .after = 5, .complete = TRUE),
    comm_sentiment_roll_51 = slider::slide_dbl(avg_comm_sentiment,
      ~ {v <- mean(.x, na.rm = TRUE); if (is.nan(v)) NA_real_ else v},
      .before = 25, .after = 25, .complete = TRUE)
  ) |>
  ungroup()
```

Let's plot one book, as a check.

```{r results=FALSE}
long_df_roll |>
  filter(bookId == "11") |>
  ggplot(aes(x = new_index, y = par_sentiment_roll_51, 
             group = source, color = source)) +
  geom_line() +
  #geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Paragraph Sentiment by Source",
    x = "Paragraph Index",
    y = "Paragraph Sentiment Score",
    color = "Source"
  ) +
  theme_minimal()
```

Now let's plot the comparison between paragraph sentiment and comment sentiment for one book, for both sources.

```{r results=FALSE}
# 1. Pivot longer so both variables are in a single 'value' column
df_long_51 <- long_df_roll |>
  pivot_longer(
    cols = c(par_sentiment_roll_51, comm_sentiment_roll_51),
    names_to = "sentiment_source",
    values_to = "sentiment_value"
  ) |>
  mutate(
    source = as.factor(source),
    # Combine source and variable for grouping
    source_var = paste(source, sentiment_source, sep = "_")
  )
```

```{r}
# 2. Define colors: same base color for each source, but with different shades
source_levels <- unique(df_long_51$source)
color_palette <- c(
  "Qidian_par_sentiment_roll_51" = "#1f77b4",
  "Qidian_comm_sentiment_roll_51"       = "#aec7e8",
  "WebNovel_par_sentiment_roll_51" = "#ff7f0e",
  "WebNovel_comm_sentiment_roll_51"       = "#ffbb78")
```

```{r}
# 3. Plot
df_long_51 |>
    filter(bookId == "11") |> #just checking a random book
ggplot(aes(x = percentage_index, y = sentiment_value, color = source_var)) +
  geom_line() +
  scale_color_manual(values = color_palette) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Paragraph and Comment Sentiment by Source",
    x = "Story progression",
    y = "Sentiment Score",
    color = "Source & Variable"
  ) +
  theme_minimal() + 
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  guides(
    color = guide_legend(nrow = 2),
    fill = guide_legend(nrow = 2)
  )
```

Both stories and comments seem to have more positive sentiment on WebNovel.

```{r}
# save the plot
ggsave("figures/sentiment_comparison_example.png", width = 8, height = 5)
```

Now let's move to the statistical modeling.

```{r message=FALSE}
library(mgcv) # for the GAM model
library(itsadug) # for visualization the estimated effects from the GAM model
library(plotfunctions)
```

Let's start with the simplest model. This is equivalent to a linear regression, because we're not adding any smooth.

```{r}
mod_GAM0 <- bam(comm_sentiment_roll_11 ~ par_sentiment_roll_11,
          data = long_df_roll, method = "ML")

summary(mod_GAM0)
```

`paragraph_sentiment` is strongly associated with `comment_sentiment`, explaining 1.05% of its variance (R-sq = 0.010).

Now let's try with a nonlinear fitting.

```{r}
mod_GAM <- gam(comm_sentiment_roll_11 ~ s(par_sentiment_roll_11),
          data = long_df_roll, method = "ML")

summary(mod_GAM)
```

The R-sq improved (0.014), so the nonlinear fitting is able to better explain the association between the two variables.

Now let's test the hypothesis that the progression of `paragraph_sentiment` matters to predict `comment_sentiment`. We start by including a second nonlinear predictor.

```{r}
mod_GAM2 <- gam(comm_sentiment_roll_11 ~ s(par_sentiment_roll_11) + s(new_index),
          data = long_df_roll, method = "ML")

summary(mod_GAM2)
```

The model has a higher R-sq (0.016), so this tells us that the progression of the story (`new_index`) goes in parallel with significantly different values of `comment_sentiment`, but we don't know if this variation is actaully related to the variation of `paragraph_sentiment`. Now let's test the interaction between `paragraph_sentiment` and the progression of the story (`new_index`). Since we know that there might be a shift in sentiment arcs due to the rolling averaging, better to smooth the nonlinear fitting usign `te()` rather than `s()` to stay more flexible ad have different smoothness in each direction (anisotropic smooths). Note that `te()` tests for both the main effect of the two predictors and their interaction, so there is no need to also include them as fixed effects outside the smooth.

```{r}
mod_GAM3 <- gam(comm_sentiment_roll_11 ~ te(par_sentiment_roll_11, new_index),
                data = long_df_roll, method = "ML")

summary(mod_GAM3)
```

R-sq (0.016). So far we haven't taken into account the fact that the data are coming from two different platforms, with two different userbases. We need to include `source` in the model to be more accurate. Since we want to allow different intercepts by `source`, we need to include it as a fixed effect as well.

```{r}
long_df_roll$source <- as.factor(long_df_roll$source)

mod_GAM3_source <- gam(comm_sentiment_roll_11 ~ source 
                       + te(par_sentiment_roll_11, new_index, by = source),
                data = long_df_roll, method = "ML")

summary(mod_GAM3_source)
```

The R-sq went from 0.01 to 0.05, so not including `source` was partly hiding the association between `pargraph_sentiment` and `comment_sentiment` because there is a lot of variability in the data across the two platforms. Let's quantify the difference in model fit.

```{r}
compareML(mod_GAM3, mod_GAM3_source) 
```

`mod_GAM3_source` has a significantly lower AIC score, which indicates a better trade-off between fit and complexity, meaning the model is more likely to generalize well to new data.

Let's see if we can fit a better model by adding a random intercept for each book, since the `comment_sentiment` for each book can be very different, depending on the plot.

```{r}
mod_GAM3_source_re <- gam(comm_sentiment_roll_11 ~ source 
                          + te(par_sentiment_roll_11, new_index, by = source)
                          + s(bookId, bs = "re"),
                          data = long_df_roll, method = "ML")

summary(mod_GAM3_source_re)
```

```{r}
compareML(mod_GAM3_source, mod_GAM3_source_re) 
```

Allowing for baseline differences in `comment_sentiment` across books further improves the explanatory power of the statistical model (significant difference in AIC), but not that much. It seems that similarities of association between `paragraph_sentiment` and `comment_sentiment` for books within each platform are stronger than individual differences between books.
